Lattices of acyclic pipe dreams

### 2023-10-30 acyclicFacets/draft/latticePipeDreams.tex
last edits before submission

\vspace*{-.8cm}

@@

@@

We recover the connection between the weak order and the Tamari lattice by the sylvester congruence for a well-chosen exiting permutation~$\omega$.
Note that \cref{thm:A} is the correct generalization of the Tamari lattice, as neither the increasing flip poset on all pipe dreams, nor its subposet induced by acyclic pipe dreams, are lattices in general (see \cref{rem:increasingFlipPosetNotLattice,rem:acyclicIncreasingFlipPosetNotLattice}).

@@

\begin{proof}
Let~$r$ be the row and $c$ be the column of the crossing~$x$. % (careful that all indices range from~$0$ to~$n$).
Since $i$ has an elbow southwest of~$x$, we have $r \le i-1$.
Since $j$ has an elbow northeast of~$x$, there are at most $r-1 \le i-2$ pipes that cross north of $(c,r)$ in column~$c$.
Hence, at least two pipes smaller than~$i$ reach the northern border weakly before column~$c$.
Excluding the pipe~$0$, there is at least one pipe $h > 0$ such that $h < i < j$ and $\omega^{-1}(h) < \omega^{-1}(j) < \omega^{-1}(i)$.
This is a $132$ pattern in the permutation~$\omega$, therefore $\omega$ is not dominant.
\end{proof}

@@

If~$m = 2$, then the pipes~$i_1$ and~$i_2$ must cross at some point~$x$.
Without loss of generality, we can assume $i_1<i_2$. 
The fact $i_1 \to i_2$ implies that $i_1$ has an elbow southwest of~$x$, while $i_2 \to i_1$ implies that $i_2$ has an elbow northeast of~$x$, so that $\omega$ is not dominant by \cref{lem:notDominant}.

>>>>

\vspace*{-1cm}

@@

\vspace*{-.5cm}

@@

We recover the connection between the weak order and the Tamari lattice by the sylvester congruence for a well-chosen exiting permutation~$\omega$.
Note that \cref{thm:A} is the correct generalization of the Tamari lattice, as neither the increasing flip poset on all pipe dreams, nor its subposet induced by acyclic pipe dreams, are lattices in general (see \cref{rem:increasingFlipPosetNotLattice,rem:acyclicIncreasingFlipPosetNotLattice}).
Other conjectural lattice structures on pipe dreams are considered in~\cite[Conj.~2.8]{Rubey}.

@@

\begin{proof}
Let~$r$ be the row and $c$ be the column of the crossing~$x$. % (careful that all indices range from~$0$ to~$n$).
Since $i$ has an elbow southwest of~$x$, we have $r \le i-1$.
Since $j$ has an elbow northeast of~$x$, there are at most $r-1 \le i-2$ pipes that cross north of $x$ in column~$c$.
Hence, at least two pipes smaller than~$i$ reach the northern border weakly before column~$c$.
Excluding the pipe~$0$, there is at least one pipe $h > 0$ such that $h < i < j$ and $\omega^{-1}(h) < \omega^{-1}(j) < \omega^{-1}(i)$.
This is a $132$ pattern in the permutation~$\omega$, therefore $\omega$ is not dominant.
\end{proof}

@@

If~$m = 2$, then the pipes~$i_1$ and~$i_2$ must cross at some point~$x$.
Without loss of generality, we can assume $i_1<i_2$. 
The fact $i_1 \to i_2$ implies that $i_1$ has an elbow southwest of~$x$, while the fact that $i_2 \to i_1$ implies that $i_2$ has an elbow northeast of~$x$, so that $\omega$ is not dominant by \cref{lem:notDominant}.
